---
title: "R_assignment"
output: html_document
---

```{r}
library(tidyverse)
library(tidyr)
library(dplyr)
```

```{r}
fang_genotypes <-read.table("fang_et_al_genotypes.txt")
snp_position <- read.table("snp_position.txt", header = TRUE, sep = "\t", fill = TRUE)
```

No.of rows and no. of columns in fang_genotypes.txt - 2783, and 986 No.of rows and no. of columns in snp_position.txt - 983 and 15

#Transposing

```{r}
Tr_fang_genotypes <- read.table("transposed_genotypes.txt")
```

#Trimming_Tr_fang_genotypes_file

```{r}
Trim_Tr_fang_data <- row_to_names(Tr_fang_genotypes, 2, remove_row = TRUE, remove_rows_above = TRUE)
```

#merging both files

```{r}
merged_data_file <- merge(snp_position, Trim_Tr_fang_data, by.x = "SNP_ID" ,by.y= "Group", all = TRUE)
# Save the merged file
write.table(merged_data_file, "merged_file.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```

#removing unwanted columns

```{r}
# Read the text file (assuming tab-separated format)
data <- read.table("merged_file.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
# Remove columns 2, 3, and 5 to 15
data_cleaned <- data[, -c(2, 5:15)]
# Save the cleaned file
write.table(data_cleaned, "cleaned_file.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```

#Maize and Teosinte files_pulled out from cleaned_file.txt

```{r}
# Load necessary library
library(dplyr)
# Read the merged file (assuming tab-separated format)
data <- read.table("cleaned_file.txt.", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
# Extract metadata columns (SNP_ID, Chromosome, Position)
metadata_cols <- data %>% select(SNP_ID, Chromosome, Position)
# Select maize genotype columns
maize_cols <- c("ZMMIL", "ZMMLR", "ZMMMR")
maize_data <- data %>% select(all_of(c("SNP_ID", "Chromosome", "Position", maize_cols)))
# Select teosinte genotype columns
teosinte_cols <- c("ZMPBA", "ZMPIL", "ZMPJA")
teosinte_data <- data %>% select(all_of(c("SNP_ID", "Chromosome", "Position", teosinte_cols)))
# Save the extracted files
write.table(maize_data, "maize_data.txt", sep = "\t", row.names = FALSE, quote = FALSE)
write.table(teosinte_data, "teosinte_data.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```

#conerting ?/? to ? symbol in maize data and teosinte data

```{r}
# Read the file (assuming tab-separated format)
data <- read.table("maize_data.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE, check.names = FALSE)
data <- read.table("teosinte_data.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE, check.names = FALSE)

# Replace "?/?" with "?" only in character columns
data[] <- lapply(data, function(x) {
  if (is.character(x)) {
    gsub("\\?/\\?", "?", x)  # Replace ?/? with ?
  } else {
    x  # Keep numeric columns unchanged
  }
})

# Save the cleaned file (ensuring tab separation is maintained)
write.table(data, "maize_for_asce.txt", sep = "\t", row.names = FALSE, quote = FALSE, col.names = TRUE)
write.table(data, "teosinte_for_asce.txt", sep = "\t", row.names = FALSE, quote = FALSE, col.names = TRUE)
```

# Ascending order of Maize chromosomes and Teosinte chromosomes

```{r}
# Load required library
library(dplyr)

# Read the maize and teosinte data files (assuming tab-separated format)
maize_data <- read.table("maize_for_asce.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
teosinte_data <- read.table("teosinte_for_asce.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# Ensure Chromosome and Position columns are numeric
maize_data$Chromosome <- as.numeric(maize_data$Chromosome)
maize_data$Position <- as.numeric(maize_data$Position)
teosinte_data$Chromosome <- as.numeric(teosinte_data$Chromosome)
teosinte_data$Position <- as.numeric(teosinte_data$Position)

# Loop through chromosomes 1 to 10 for both maize and teosinte
for (chr in 1:10) {
  # Process Maize Data
  maize_chr_data <- maize_data %>% filter(Chromosome == chr)
  maize_chr_sorted <- maize_chr_data %>% arrange(Position)
  write.table(maize_chr_sorted, paste0("maize_chr", chr, "_ascending.txt"), sep = "\t", row.names = FALSE, quote = FALSE)

  # Process Teosinte Data
  teosinte_chr_data <- teosinte_data %>% filter(Chromosome == chr)
  teosinte_chr_sorted <- teosinte_chr_data %>% arrange(Position)
  write.table(teosinte_chr_sorted, paste0("teosinte_chr", chr, "_ascending.txt"), sep = "\t", row.names = FALSE, quote = FALSE)
}
```

#Creating new files with the conversion of ? into - for decresing order in maize and teosinte data

```{r}
# Function to replace "?" with "-" and save as a new file
replace_question_marks <- function(input_file, output_file) {
  # Read the file (assuming tab-separated format)
  data <- read.table(input_file, header = TRUE, sep = "\t", stringsAsFactors = FALSE, check.names = FALSE)
  
  # Replace "?" with "-" only in character columns
  data[] <- lapply(data, function(x) {
    if (is.character(x)) {
      gsub("\\?", "-", x)  # Replace ? with -
    } else {
      x  # Keep numeric columns unchanged
    }
  })
  
  # Save the converted file with a new name
  write.table(data, output_file, sep = "\t", row.names = FALSE, quote = FALSE, col.names = TRUE)
}

# Process Maize Files
replace_question_marks("maize_for_asce.txt", "maize_for_desce.txt")

# Process Teosinte Files
replace_question_marks("teosinte_for_asce.txt", "teosinte_for_desce.txt")
```

Descending order of Maize and Teosinte chromsomes

```{r}
# Load required library
library(dplyr)

# Read the maize and teosinte data files (assuming tab-separated format)
maize_data <- read.table("maize_for_desce.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
teosinte_data <- read.table("teosinte_for_desce.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# Ensure Chromosome and Position columns are numeric
maize_data$Chromosome <- as.numeric(maize_data$Chromosome)
maize_data$Position <- as.numeric(maize_data$Position)
teosinte_data$Chromosome <- as.numeric(teosinte_data$Chromosome)
teosinte_data$Position <- as.numeric(teosinte_data$Position)

# Loop through chromosomes 1 to 10 for both maize and teosinte
for (chr in 1:10) {
  # Process Maize Data
  maize_chr_data <- maize_data %>% filter(Chromosome == chr)
  maize_chr_sorted <- maize_chr_data %>% arrange(desc(Position))
  write.table(maize_chr_sorted, paste0("maize_chr", chr, "_descending.txt"), sep = "\t", row.names = FALSE, quote = FALSE)

  # Process Teosinte Data
  teosinte_chr_data <- teosinte_data %>% filter(Chromosome == chr)
  teosinte_chr_sorted <- teosinte_chr_data %>% arrange(desc(Position))
  write.table(teosinte_chr_sorted, paste0("teosinte_chr", chr, "_descending.txt"), sep = "\t", row.names = FALSE, quote = FALSE)
}
```

Part-II Visualization

```{r}
# Load libraries
library(ggplot2)
library(dplyr)
library(tidyr)
```
Checking again

```{r}
snp_data <- read.table("teosinte_data.txt", header = TRUE, sep = "\t")
snp_data <- read.table("teosinte_data.txt", header = TRUE, sep = "\t", fill = TRUE)
lines <- readLines("teosinte_data.txt")
lengths <- sapply(strsplit(lines, "\t"), length)
table(lengths)
snp_data <- read.table("teosinte_data.txt", header = TRUE, sep = "\t", fill = TRUE, comment.char = "")
snp_data <- read.table("teosinte_data.txt", sep = "\t", fill = TRUE, header = FALSE)
colnames(snp_data) <- c("Column1", "Column2", "Column3", "Column4", "Column5", "Column6")  # Adjust as needed

```
#Merging Maize and Teosinte data sets for visualization

```{r}
# Load necessary library
library(dplyr)

# Read maize data
maize_data <- read.table("maize_data.txt", header = TRUE, sep = "\t", fill = TRUE, stringsAsFactors = FALSE)

# Read teosinte data
teosinte_data <- read.table("teosinte_data.txt", header = TRUE, sep = "\t", fill = TRUE, stringsAsFactors = FALSE)

# Check if both datasets have the same number of columns
if (ncol(maize_data) != ncol(teosinte_data)) {
  stop("Error: Maize and Teosinte datasets have different column numbers. Check your files!")
}

# Merge the two datasets
combined_data <- bind_rows(maize_data, teosinte_data)

# Display structure of the merged data
str(combined_data)

# Save the merged dataset
write.table(combined_data, "merged_snp_data.txt", sep = "\t", row.names = FALSE, quote = FALSE)

# View first few rows
head(combined_data)

```
# original_visualization_part

```{r}
# Load required libraries
library(dplyr)
library(tidyr)
library(ggplot2)

# Read Maize and Teosinte SNP Data
maize_data <- read.table("maize_data.txt", header = TRUE, sep = "\t", fill = TRUE, stringsAsFactors = FALSE)
teosinte_data <- read.table("teosinte_data.txt", header = TRUE, sep = "\t", fill = TRUE, stringsAsFactors = FALSE)

# Ensure both datasets have the same column names
colnames(teosinte_data) <- colnames(maize_data)

# Merge datasets & add a new column to distinguish groups
maize_data$Group <- "Maize"
teosinte_data$Group <- "Teosinte"

combined_data <- bind_rows(maize_data, teosinte_data)

# Convert Chromosome column to factor with correct numerical order
combined_data$Chromosome <- factor(combined_data$Chromosome, 
                                   levels = as.character(1:10), 
                                   ordered = TRUE)

# ---------------------------------
# 1️⃣ SNP Count per Chromosome (Correct Order)
# ---------------------------------

# Count SNPs per chromosome for each group
snp_count <- combined_data %>%
  group_by(Chromosome, Group) %>%
  summarise(SNP_Count = n(), .groups = 'drop')

# Plot SNP distribution across chromosomes (ordered correctly)
ggplot(snp_count, aes(x = Chromosome, y = SNP_Count, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "SNP Distribution Across Chromosomes",
       x = "Chromosome", y = "SNP Count") +
  theme_minimal()

# ---------------------------------
# 2️⃣ Missing Data & Heterozygosity
# ---------------------------------

# Define heterozygous and missing data
combined_data <- combined_data %>%
  pivot_longer(cols = -c(Chromosome, Position, Group), names_to = "Sample", values_to = "Genotype") %>%
  mutate(
    Heterozygosity = case_when(
      Genotype %in% c("A/A", "C/C", "G/G", "T/T") ~ "Homozygous",
      Genotype %in% c("A/T", "C/G", "A/C", "G/T") ~ "Heterozygous",
      Genotype == "NA" ~ "Missing",
      TRUE ~ "Unknown"
    )
  )

# Ensure Chromosome order remains correct
combined_data$Chromosome <- factor(combined_data$Chromosome, 
                                   levels = as.character(1:10), 
                                   ordered = TRUE)

# Calculate proportions
het_summary <- combined_data %>%
  group_by(Group, Heterozygosity) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  mutate(Proportion = Count / sum(Count))

# Plot proportion of homozygous, heterozygous, and missing data
ggplot(het_summary, aes(x = Group, y = Proportion, fill = Heterozygosity)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Proportion of Homozygous, Heterozygous, and Missing Data",
       x = "Group", y = "Proportion") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

# ---------------------------------
# 3️⃣ SNP Density Along Chromosomes (Correct Order)
# ---------------------------------

# SNP density visualization using density plot
ggplot(combined_data, aes(x = Position, fill = Group)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ Chromosome, scales = "free_x") +
  labs(title = "SNP Density Along Chromosomes",
       x = "Genomic Position", y = "Density") +
  theme_minimal()

```











